---
title: "Summarising and Exploring Data"
output: html_document
date: '2022-12-11'
editor_options: 
  chunk_output_type: console
---

## Week 2

## Summarising and Data Exploration

Importing the data- Data was already downloaded, imported the data directly. 

Loading libraries
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(ggpubr)
```

Summary of the data- The summary of n.cpg and t.cpg are only taken since these were already filtered. 
```{r warning=FALSE, message=FALSE}
summary(n.cpg)

summary(t.cpg)
```

Merging datasets
```{r warning=FALSE, message=FALSE}
merged <- bind_rows(n.cpg=n.cpg, t.cpg=t.cpg, .id = "id")

colnames(merged)[1]="COLO829"
```

Removing random chromosomes in merged dataset
```{r warning=FALSE, message=FALSE}
final <- subset(merged, !(endsWith(V1, '_random') | endsWith(V1, 'v1') | endsWith(V1, 'v2') | endsWith(V1, 'BV')))
```

Density Plots of 4 datasets
1. COLO829_1
```{r warning=FALSE, message=FALSE}
plot(density(colo829_1_filter$V11, na.rm = TRUE, xlab="Frequency"))
```

2. COLO829BL_1 
```{r warning=FALSE, message=FALSE}
plot(density(colo829bl_1_filter$frequency, na.rm = TRUE))
```

3. n.cpg 
```{r warning=FALSE, message=FALSE}
plot(density(n.cpg$V11, na.rm = TRUE), xlab="Frequency", main="Density plot of Frequency of Normal sample")
```

4. t.cpg
```{r warning=FALSE, message=FALSE}
plot(density(t.cpg$V11, na.rm = TRUE), xlab="Frequency", main="Density plot of Frequency of Tumour sample")
```

Density plot of t.cpg and n.cpg showing different strands on the same plot
1. t.cpg
```{r warning=FALSE, message=FALSE}
ggplot(t.cpg, aes(x=V11, fill=V6, color=V6))+
  geom_density(alpha=0.2)+
  labs(x="Frequency", y="Strand", title = "Density plot of frequency showing strands")
```

2. n.cpg
```{r warning=FALSE, message=FALSE}
ggplot(n.cpg, aes(x=V11, fill=V6, color=V6))+
  geom_density(alpha=0.2)+
  labs(x="Frequency", y="Strand", title = "Density plot of frequency showing strands")
```

Density Plot of both the samples in the same graph segregated by forward or reverse strand
1. Forward Strand
```{r warning=FALSE, message=FALSE}
strand_1 <- final %>% 
  filter(V6=="+")

plot_strand_1 <- strand_1 %>% 
  ggplot(aes(x=V11, colour=COLO829))+
  geom_density(alpha=0.2)+
  labs(x="Methylation Frequency (%)", title = "Density plot of Methylation Frequency of Forward strand")
```

2. Reverse Strand
```{r warning=FALSE, message=FALSE}
strand_2 <- final %>% 
  filter(V6=="-")

plot_strand_2 <- strand_2 %>% 
  ggplot(aes(x=V11, colour=COLO829))+
  geom_density(alpha=0.2)+
  labs(x="Methylation Frequency (%)", title = "Density plot of Methylation Frequency of Reverse Strand")
```

Box plot of t.cpg and n.cpg datasets of different strands in the same plot
1. t.cpg
```{r warning=FALSE, message=FALSE}
ggplot(t.cpg, aes(x=V6, y=V11, fill=V6))+
  geom_boxplot()+
  labs(x="Strand", y="Frequency", title="Plot of frequency with respect to strand")+
  theme_classic()
```

2. n.cpg
```{r warning=FALSE, message=FALSE}
ggplot(n.cpg, aes(x=V6, y=V11, fill=V6))+
  geom_boxplot()+
  labs(x="Strand", y="Frequency", title="Plot of frequency with respect to strand")+
  theme_classic()
```

Box plot of different chromosomes
1. t.cpg
```{r warning=FALSE, message=FALSE}
ggplot(t_cpg_filter, aes(x=seq_name, y=frequency))+geom_boxplot()
```

2. n.cpg
```{r warning=FALSE, message=FALSE}
ggplot(n.cpg, aes(x=V1, y=V11, fill=V1))+
  geom_boxplot()+
  labs(x="Chromosome", y="Frequency", title = "Plot of chromosome and their frequency")+
  theme(axis.text.x = element_text(angle=90, hjust = 1))+
  coord_flip()
```

Boxplots of different chromosomes
```{r warning=FALSE, message=FALSE}
subset_1 <- final %>% 
  filter(V1=="chr21")

subset_2 <- final %>% 
  filter(V1=="chr22")

subset_3 <- final %>% 
  filter(V1=="chrX")

subset_4 <- final %>% 
  filter(V1=="chrY")

comparison=list(c("n.cpg", "t.cpg"))

plot_1 <- ggplot(subset_1, aes(x=COLO829, y=V11, fill=COLO829))+
  geom_boxplot()+
  labs(x="Sample", y="Methylation Frequency (%)", title = "Chromosome21")+
  stat_compare_means(comparisons = comparison, paired = FALSE, method = "wilcox", label = "p.format")+stat_boxplot()

plot_2 <- ggplot(subset_2, aes(x=COLO829, y=V11, fill=COLO829))+
  geom_boxplot()+
  labs(x="Sample", y="Methylation Frequency (%)", title = "Chromosome22")+
  stat_compare_means(comparisons = comparison, paired = FALSE, method = "wilcox", label = "p.format")+stat_boxplot()

plot_3 <- ggplot(subset_3, aes(x=COLO829, y=V11, fill=COLO829))+
  geom_boxplot()+
  labs(x="Sample",y="Methylation Frequency (%)", title = "ChromosomeX")+
  stat_compare_means(comparisons = comparison, paired = FALSE, method = "wilcox", label = "p.format")+stat_boxplot()

plot_4 <- ggplot(subset_4, aes(x=COLO829, y=V11, fill=COLO829))+
  geom_boxplot()+
  labs(x="Sample", y="Methylation Frequency (%)", title = "ChromosomeY")+
  stat_compare_means(comparisons = comparison, paired = FALSE, method = "wilcox", label = "p.format")+stat_boxplot()

plot <- ggarrange(plot_3, plot_4, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom")
annotate_figure(plot, top = text_grob("Boxplot plot of COLO829", face = "bold", size = 14))

```

Boxplot for Forward strand of both samples in the same plot
```{r warning=FALSE, message=FALSE}
strand_1 <- final %>% 
  filter(V6=="+")

comparison=list(c("n.cpg", "t.cpg"))

strand_box_1 <- strand_1 %>% 
  ggplot(aes(x=COLO829, y=V11, fill=COLO829))+
  geom_boxplot()+
  labs(x="Sample", y="Methylation Frequency (%)", title="Plot of Methylation Frequency of Forward Strand")+
  stat_compare_means(comparisons = comparison, paired = FALSE, method = "wilcox", label = "p.format")+stat_boxplot()
```

Boxplot for Reverse strand of both samples in the same plot
```{r warning=FALSE, message=FALSE}
strand_2 <- final %>% 
  filter(V6=="-")

comparison=list(c("n.cpg", "t.cpg"))

strand_box_2 <- strand_2 %>% 
  ggplot(aes(x=COLO829, y=V11, fill=COLO829))+
  geom_boxplot()+
  labs(x="Sample", y="Methylation Frequency (%)", title="Plot of Methylation Frequency of Reverse Strand")+
  stat_compare_means(comparisons = comparison, paired = FALSE, method = "wilcox", label = "p.format")+stat_boxplot()
```

Wilcox Test on each chromosome
```{r warning=FALSE, message=FALSE}
wilcox_dt=data.frame()

for (i in unique(final$V1)){
  xi= final[which(final[, "V1"] %in% i), ]
  xi_n=xi[which(xi$COLO829 %in% "n.cpg"), ]
  xi_t=xi[which(xi$COLO829 %in% "t.cpg"), ]

  wilx_i=wilcox.test(xi_t$V11, xi_n$V11)
  mean_diff_i=mean(xi_t$V11, na.rm=TRUE)-mean(xi_n$V11, na.rm=TRUE)
  out=rbind(c(i, round(mean_diff_i,5), wilx_i$p.value))
  
  wilcox_dt=rbind.data.frame(out, wilcox_dt)
  
  print(i)
  
}

colnames(wilcox_dt)=c("chr", "mean_diff", "p.value")

all.chr=paste0("chr", c(1:22, "X", "Y"))

wilcox_dt=wilcox_dt[wilcox_dt$chr %in% all.chr, ]
wilcox_dt$mean_diff=as.numeric(wilcox_dt$mean_diff)
wilcox_dt$p.value=as.numeric(wilcox_dt$p.value)

wilcox_dt$adjusted.pvalue=p.adjust(wilcox_dt$p.value, method = "BH")
```

Density Plots of each chromosome of the samples in the same plot
```{r warning=FALSE, message=FALSE}
plot_1 <- ggplot(subset_1, aes(x=V11, colour=COLO829))+
  geom_density()+
  labs(x="Methylation Frequency (%)", title = "ChromosomeM")

plot_2 <- ggplot(subset_2, aes(x=V11, colour=COLO829))+
  geom_density()+
  labs(x="Methylation Frequency (%)", title = "Chromosome22")

plot_3 <-ggplot(subset_3, aes(x=V11, colour=COLO829))+
  geom_density()+
  labs(x="Methylation Frequency (%)", title = "ChromosomeX")

plot_4 <- ggplot(subset_4, aes(x=V11, colour=COLO829))+
  geom_density()+
  labs(x="Methylation Frequency (%)", title = "ChromosomeY")

plot <- ggarrange(plot_1, plot_2, plot_3, plot_4, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")
annotate_figure(plot, top = text_grob("Density plot of COLO829", face = "bold", size = 14))
```




